---
title: "DBS matrix maker"
author: "Ellie Dunstone"
date: "2023-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include=FALSE}
library(tidyverse)
library(grid)
library(gtable)
```

## Introduction

This is a script for generating a DBS78 matrix (see <https://cancer.sanger.ac.uk/signatures/dbs/>) for signature analysis from a table of DNVs. This is based on a script written by Andrew Lawson.

You can generate the table of DNVs using the script "dbs_calling_annotating_runs_subs.R".

## Read in and format data

Read in your table of DNVs.

```{r read DNV table}
dataset <- read.table(file = "/Users/ed4/Documents/phd/chemo_project/BotSeq/panbody/panbody_20230106/data/dbs_calling/dbs_calls.csv", sep = ",", stringsAsFactors = F, header = T)
```

Filter for DBS only (should already be the case if generated using "dbs_calling_annotating_runs_subs.R").

```{r format table}
#check/filter for DBS only
dnv_calls <- dataset[which(str_length(dataset$ref) == 2 & str_length(dataset$mut) == 2),]
```

Create an array of the 78 DBS mutation types.

```{r init DBS types}
dnv_types <- array(data = NA,dim = c(78,2))

dnv_types[,1] <- c("AC>CA","AC>CG","AC>CT","AC>GA","AC>GG","AC>GT","AC>TA","AC>TG","AC>TT",
                   "AT>CA","AT>CC","AT>CG","AT>GA","AT>GC","AT>TA",
                   "CC>AA","CC>AG","CC>AT","CC>GA","CC>GG","CC>GT","CC>TA","CC>TG","CC>TT",
                   "CG>AT","CG>GC","CG>GT","CG>TA","CG>TC","CG>TT",
                   "CT>AA","CT>AC","CT>AG","CT>GA","CT>GC","CT>GG","CT>TA","CT>TC","CT>TG",
                   "GC>AA","GC>AG","GC>AT","GC>CA","GC>CG","GC>TA",
                   "TA>AT","TA>CG","TA>CT","TA>GC","TA>GG","TA>GT",
                   "TC>AA","TC>AG","TC>AT","TC>CA","TC>CG","TC>CT","TC>GA","TC>GG","TC>GT",
                   "TG>AA","TG>AC","TG>AT","TG>CA","TG>CC","TG>CT","TG>GA","TG>GC","TG>GT",
                   "TT>AA","TT>AC","TT>AG","TT>CA","TT>CC","TT>CG","TT>GA","TT>GC","TT>GG")

```

Assign mutation types to dnv calls table.

```{r assign mut types}

#generate complementation table
comp_table <- array(data = NA, dim = c(4,2))

comp_table[,1] <- c("A","C","G","T") 
comp_table[,2] <- c("T","G","C","A") 


for(i in 1:nrow(dnv_types)) {
  dnv_types[i,2] <- paste(comp_table[which(comp_table[,1] == substr(dnv_types[i,1],2,2)),2],
                          comp_table[which(comp_table[,1] == substr(dnv_types[i,1],1,1)),2],
                          ">",
                          comp_table[which(comp_table[,1] == substr(dnv_types[i,1],5,5)),2],
                          comp_table[which(comp_table[,1] == substr(dnv_types[i,1],4,4)),2],
                          sep = "")
}

#assign mut types to dnv table
dnv_calls$mut_type <- paste(dnv_calls$ref,">",dnv_calls$mut,sep = "")
dnv_calls$dnv_mut <- dnv_calls$mut_type
```

Complement mutations where necessary.

```{r complement}
#use complementation table to make mutation types strand-agnostic
for(i in 1:nrow(dnv_calls)) {
  if(dnv_calls$dnv_mut[i] %in% dnv_types[,2]) {
    dnv_calls$dnv_mut[i] <- dnv_types[which(dnv_types[,2] == dnv_calls$dnv_mut[i]),1]
  }
}
```

## Plot and generate output

We can now generate plots and outputs.

```{r spectra plot}

samples <- unique(dnv_calls$sampleID) #generate sample list

for(k in 1:length(samples)) {
  dnv_sample_table <- dnv_calls[which(dnv_calls$sampleID == samples[k]),]
  
  dnv_counts <- as.data.frame(array(data = NA, dim = c(78,4)))
  colnames(dnv_counts) <- c("dnv_type","dnv_count","dnv_sub","dnv_alt")
  
  dnv_counts[,1] <- dnv_types[,1] 
  
  for(i in 1:nrow(dnv_counts)) {
    dnv_counts[i,2] <- length(which(dnv_sample_table$dnv_mut == dnv_counts[i,1]))
    dnv_counts[i,3] <- paste(substr(dnv_counts[i,1],1,3),"NN",sep = "")
    dnv_counts[i,4] <- substr(dnv_counts[i,1],4,5)
  }
  
  
  color_palette <- c("#A7CEE2","#2679B2","#B3DE8E","#399F34",
                     "#F99B9B","#E02427","#FCBE75","#FD7F23",
                     "#CAB3D5","#694098")
  
  
  dnv_plot <- ggplot(dnv_counts, aes(x = dnv_type, y = dnv_count)) +
    geom_bar(stat = "identity", aes(fill = dnv_sub), width = 0.5) +
    scale_fill_manual(values = color_palette) +
    scale_x_discrete(labels = dnv_counts$dnv_alt) +
    scale_y_continuous(expand = c(0,0), limits = c(0,max(dnv_counts$dnv_count) + max(dnv_counts$dnv_count) / 10), breaks = c(seq(from = 0,to = ceiling(max(dnv_counts$dnv_count) + max(dnv_counts$dnv_count) / 10),by = ceiling(ceiling(max(dnv_counts$dnv_count) + max(dnv_counts$dnv_count) / 10) / 10)))) +
    theme_bw() +
    guides(fill = "none") +
    theme(axis.ticks = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          axis.text.x = element_text(angle = 90, size = 6, vjust = 0.5),
          axis.title = element_blank())
  
  dnv_grob_plot <- ggplotGrob(dnv_plot)
  panel_id <- dnv_grob_plot$layout[dnv_grob_plot$layout$name == "panel",c("t","l")]
  dnv_grob_plot <- gtable_add_rows(dnv_grob_plot, unit(1.5, "cm"), 0)
  
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 0.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#A7CEE2", col = "white")),t = 1, l = panel_id$l, name = "ac_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 9.2 / 78, y = 0, hjust = 0, vjust = 0, width = 5.8 / 78, height = 0.5, gp=gpar(fill = "#2679B2", col = "white")),t = 1, l = panel_id$l, name = "at_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 15.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#B3DE8E", col = "white")),t = 1, l = panel_id$l, name = "cc_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 24.2 / 78, y = 0, hjust = 0, vjust = 0, width = 5.8 / 78, height = 0.5, gp=gpar(fill = "#399F34", col = "white")),t = 1, l = panel_id$l, name = "cg_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 30.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#F99B9B", col = "white")),t = 1, l = panel_id$l, name = "ct_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 39.2 / 78, y = 0, hjust = 0, vjust = 0, width = 5.8 / 78, height = 0.5, gp=gpar(fill = "#E02427", col = "white")),t = 1, l = panel_id$l, name = "gc_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 45.2 / 78, y = 0, hjust = 0, vjust = 0, width = 5.8 / 78, height = 0.5, gp=gpar(fill = "#FCBE75", col = "white")),t = 1, l = panel_id$l, name = "ta_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 51.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#FD7F23", col = "white")),t = 1, l = panel_id$l, name = "tc_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 60.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#CAB3D5", col = "white")),t = 1, l = panel_id$l, name = "tg_grob")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, rectGrob(x = 69.2 / 78, y = 0, hjust = 0, vjust = 0, width = 8.8 / 78, height = 0.5, gp=gpar(fill = "#694098", col = "white")),t = 1, l = panel_id$l, name = "tt_grob")
  
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "AC>NN", x = (0.2 / 78 + 0.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "ac_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "AT>NN", x = (9.2 / 78 + 9.2 / 78 + 5.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "at_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "CC>NN", x = (15.2 / 78 + 15.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "cc_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "CG>NN", x = (24.2 / 78 + 24.2 / 78 + 5.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "cg_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "CT>NN", x = (30.2 / 78 + 30.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "ct_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "GC>NN", x = (39.2 / 78 + 39.2 / 78 + 5.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "gc_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "TA>NN", x = (45.2 / 78 + 45.2 / 78 + 5.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "ta_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "TC>NN", x = (51.2 / 78 + 51.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "tc_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "TG>NN", x = (60.2 / 78 + 60.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "tg_text")
  dnv_grob_plot <- gtable_add_grob(dnv_grob_plot, textGrob(label = "TT>NN", x = (69.2 / 78 + 69.2 / 78 + 8.8 / 78) / 2, y = 0.6, hjust = 0.5, vjust = 0),t = 1, l = panel_id$l, name = "tt_text")
  
  ggsave(filename = paste("/Users/ed4/Documents/phd/chemo_project/BotSeq/panbody/panbody_20230106/data/dbs_calling/DBS_spectra/DBS78_",samples[k],"_spectrum.pdf",sep = ""), height = 4, width = 10, plot = dnv_grob_plot)
  
}
```

Generate DBS78 matrix

```{r make matrix}
samples <- unique(dnv_calls$sampleID) #generate sample list

#init matrix
dbs_matrix <- as.data.frame(array(data = NA, dim = c(78,length(samples)+1)))
dbs_matrix[,1] <- dnv_types[,1]
colnames(dbs_matrix)[1] <- "MutationType"


for(k in 1:length(samples)) {
  dnv_sample_table <- dnv_calls[which(dnv_calls$sampleID == samples[k]),]
  
  dnv_counts <- as.data.frame(array(data = NA, dim = c(78,4)))
  colnames(dnv_counts) <- c("dnv_type","dnv_count","dnv_sub","dnv_alt")
  
  dnv_counts[,1] <- dnv_types[,1]
  colnames(dnv_counts) <- c("dnv_type","dnv_count","dnv_sub","dnv_alt")
  
  dnv_counts[,1] <- dnv_types[,1] 
  
  for(i in 1:nrow(dnv_counts)) {
    dnv_counts[i,2] <- length(which(dnv_sample_table$dnv_mut == dnv_counts[i,1]))
    dnv_counts[i,3] <- paste(substr(dnv_counts[i,1],1,3),"NN",sep = "")
    dnv_counts[i,4] <- substr(dnv_counts[i,1],4,5)
  }
  
  
  colnames(dbs_matrix)[k+1] <- samples[k]
  dbs_matrix[,k+1] <- dnv_counts$dnv_count
}
```

Write out matrix

```{r output matrix}
write.table(dbs_matrix, "/Users/ed4/Documents/phd/chemo_project/BotSeq/panbody/panbody_20230106/data/dbs_mutation_matrix.tsv", sep = "\t", row.names = FALSE)
```
